Linear vertical shear instability in astrophysical disks: a semi-numerical study

### 2014-12-19 linear_adia.tex
regular updates

We now have an equation in the same form as Eq. \ref{iso_ode3}:
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
% We again seek power-series solutions 
% \begin{align}
%   g(\zhat) = \sum_{m=0}^\infty b_m \zhat^m,
% \end{align}
% which requires
% \begin{align}
%   (m+2)(m+1)b_{m+2} + \left[\left(2\alpha - A\right)m + B + \alpha\right]=0.
% \end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).
\end{align}
Inserting the definition of $B$ we get
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
  \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
  \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}. 
\end{align}
This gives the eigenfrequency $\hat{sigma}$ and completes the solution
description.

@@

The acceptable solution for $\alpha$ when $\gamma=2$ is
\begin{align}
  \alpha = \frac{1}{2}\left[\left(1+\ii\epsilon q \hat{k}\right) -
    \sqrt{1 + \hat{k}^2\left(2-\epsilon^2 q^2\right)}\right]. 
\end{align}
Since $\epsilon\ll 1$, $\real\alpha<0$, so the vertical velocity
eventually decays for large heights. For $\hat{k}\gg 1$ the
perturbation decays rapidly away from the midplane. This is again
unlike isothermal perturbations, whose magnitude increase with $|z|$.

@@

We can also compare growth rates in the limit $\hat{k}\gg1$ using
Eq. \ref{simple_growth} and Eq. \ref{gam2_growth},
\begin{align}
  &\hat{\nu} \geq \sqrt{\frac{|\epsilon q|}{2\hat{k}}}
  &  \text{as }  \hat{k}\to\infty  \text{ for }  \gamma=1, \\
  &\hat{\nu} \leq \frac{|\epsilon q|}{2^{7/4}\sqrt{\hat{k}}}
  &\text{as }  \hat{k}\to\infty  \text{ for }  \gamma=2.
\end{align}
Then the growth rate of perturbations with large $|\hat{k}|$ in a
$\gamma=2$ disk is at most $2^{-5/4}\sqrt{|\epsilon q|}$ times that in a
$\gamma=1$ disk. For $\epsilon = 0.1$ and $|q|=1$, this factor is
$\sim 0.1$.

@@

>>>>

We now have an equation in the same form as Eq. \ref{iso_ode3}:
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
% We again seek power-series solutions 
% \begin{align}
%   g(\zhat) = \sum_{m=0}^\infty b_m \zhat^m,
% \end{align}
% which requires
% \begin{align}
%   (m+2)(m+1)b_{m+2} + \left[\left(2\alpha - A\right)m + B + \alpha\right]=0.
% \end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).
\end{align}
Inserting the definition of $B$ we get
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
  \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
  \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}. 
\end{align}
This gives the eigenfrequency $\hat{\sigma}$ and completes the solution
description.

@@

The acceptable solution for $\alpha$ when $\gamma=2$ is
\begin{align}
  \alpha = \frac{1}{2}\left[\left(1+\ii\epsilon q \hat{k}\right) -
    \sqrt{1 + \hat{k}^2\left(2-\epsilon^2 q^2\right)}\right].\label{gam2_alpha}
\end{align}
Since $\epsilon\ll 1$, $\real\alpha<0$, so the vertical velocity
eventually decays for large heights. For $\hat{k}\gg 1$ the
perturbation decays rapidly away from the midplane. This is again
unlike isothermal perturbations, whose magnitude increase with $|z|$.

@@

We can also compare growth rates in the limit $\hat{k}\gg1$ using
Eq. \ref{simple_growth} and Eq. \ref{gam2_growth},
\begin{align}
  &\hat{\nu} \geq \sqrt{\frac{|\epsilon q|}{2\hat{k}}}
  &  \text{as }  \hat{k}\to\infty  \text{ for }  \gamma=1, \\
  &\hat{\nu} \leq \frac{|\epsilon q|}{2^{7/4}\sqrt{\hat{k}}}
  &\text{as }  \hat{k}\to\infty  \text{ for }  \gamma=2.\label{gam2_growth_rate}
\end{align}
Then the growth rate of perturbations with large $|\hat{k}|$ in a
$\gamma=2$ disk is at most $2^{-5/4}\sqrt{|\epsilon q|}$ times that in a
$\gamma=1$ disk. For $\epsilon = 0.1$ and $|q|=1$, this factor is
$\sim 0.1$.

@@

\subsubsection{Qualitative expectations in the presence of thermal
  relaxation}

### 2014-12-19 numerical.tex
regular updates

\subsection{Vertically isothermal disks with isothermal perturbations}   
We first validate the low-frequency approximation by re-producing
results from \cite{mcnally14} who solved the full problem for
vertically isothermal disks ($\Gamma=1$) subject to isothermal
perturbations ($\gamma=1$). The parameters are $q=-1$,
$p=-1.5$, $\epsilon=0.1$ and $k_x = 200\pi/r$ (or $\hat{k} = 20\pi$),
with solid vertical boundaries.

For this problem we solve Eq. \ref{iso_ode1}, using a pseudo-spectral
method by expanding $W$ in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equation on a grid with
$\zmax=10H_\mathrm{iso}$. This procedure converts the linear problem
to a standard matrix eigenvalue problem.

@@

\subsection{Stabilization of the VSI by a positive vertical entropy
  gradient}

>>>>

\subsection{Vertically isothermal disks with isothermal
  perturbations}\label{vertiso_pertiso} 
We first validate the low-frequency approximation by re-producing
results from \cite{mcnally14} who solved the full problem for
vertically isothermal disks ($\Gamma=1$) subject to isothermal
perturbations ($\gamma=1$). The parameters are $q=-1$,
$p=-1.5$, $\epsilon=0.1$ and $k_x = 200\pi/r$ (or $\hat{k} = 20\pi$),
with solid vertical boundaries.

For this problem we solve Eq. \ref{iso_ode}, with $D$ replaced by
$\kappa^2$, using a pseudo-spectral
method by expanding $W$ in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equation on a grid with
$\zmax=10\epsilon r$. This procedure converts the linear problem
to a standard matrix eigenvalue problem.

@@

\subsection{Stabilization of the VSI by a positive vertical entropy
  gradient}
We demonstrate the (strong) stabilizing effect of a  
positive vertical entropy gradient. We solve the more general form of
the linear problem, Eq. \ref{ode_w}---\ref{ode_Q}, again with the
replacement $D\to\kappa^2$. We consider an effectively vertically
isothermal disk by setting $\Gamma=1.011$ with adiabatic index
$\gamma=2$. We also set $\beta=10^9$ so the cooling term is
effectively switched off. Other disk and perturbation parameters are
the same as in \S\ref{vertiso_pertiso}.

We use the same pseudo-spectral method as before. Here we set
$\zmax = 2\epsilon r$ and impose a free surface. This boundary
condition was found to be most convenient for numerical
implementation. Since we expect the perturbations to decay rapidly
away from the midplane (see \S\ref{analytic_adia}), we do not expect 
boundary conditions to play an important role, provided that $\zmax$ is
sufficiently large.

Fig. \ref{lowfreq_eigen_adia} show the eigenvalues for this
problem. The growth rates are about an order of magnitude smaller than
the previous case. The fundamental VSI mode is that with
$\mathrm{max}|\nu|$. Its expected and numerically-calculated growth
rates compares well, with 
\begin{align*}
  \nu = 0.03751 \epsilon \Omega_k &\quad \text{(from
    Eq. \ref{gam2_growth_rate})},\\
  \nu = 0.03682 \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
The fundamental mode is plotted in Fig. \ref{lowfreq_eigenfunc_adia},
and clearly show that perturbations are rapidly stabilized away from
the midplane. For the adopted disk and perturbation parameters 
($|q|=1$ and $\hat{k} = 20\pi$), Eq. \ref{gam2_alpha} gives $\real{\alpha}\simeq - 43.8$,
corresponding to a characteristic decay lengthscale of $\simeq
0.15\epsilon r$, which is consistent with that observed in 
Fig. \ref{lowfreq_eigenfunc_adia}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_adia}
  \caption{Eigenvalues in the low-frequency approximation for the
    vertical shear instability in a nearly-vertically isothermal disk
    ($\Gamma=1.011$) evolved adiabatically with $\gamma=2$ and
    $\beta=10^9$. The disk
    parameters are $q=-1$, 
    $p=-1.5$ and $\epsilon=0.1$, while the perturbation radial
    wavenumber is $k_x=200\pi/r$. \label{lowfreq_eigen_adia}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvectorvz_adia}
  \caption{Fundamental VSI mode in a nearly-vertically
    isothermal disk ($\Gamma=1.011$) evolved adiabatically
    ($\gamma=2$, $\beta=10^9$). This eigenfunction 
    corresponds to the top-left eigenvalue displayed in 
    Fig. \ref{lowfreq_eigen_adia} (largest $|\nu|$).  
    The solid and dashed lines are $\real \delta v_z$ and $\imag
    \delta v_z$, respectively.  
    \label{lowfreq_eigenfunc_adia}
  }
\end{figure}

\subsection{Effect of cooling}

### 2014-12-19 setup.tex
regular updates

We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)}
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant. 
The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1}. 
\end{align}

@@

The axisymmetric VGSB equations read 
\begin{align}
  & \frac{\p\rho}{\p t} + \frac{\p}{\p x}\left(\rho v_x\right) +
  \frac{\p}{\p z}\left(\rho v_z\right)=0,\\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
  v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}} = - 
  2\Omega_0\hat{\bm{z}}\times\bm{v} - S_0v_x\hat{\bm{y}} \notag\\ 
  &\phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
    v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}}=}-\frac{1}{\rho}\nabla P - \frac{\p\Phi_0}{\p z}\hat{\bm{z}},\\
  & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P = -\gamma P
  \nabla\cdot\bm{v} -
  \frac{\rho}{t_c}\left(\frac{P}{\rho}-\left.\frac{P}{\rho}\right|_{t=0}\right), 
\end{align}
where $\bm{v}$ is the velocity field relative to the bulk flow
(i.e. $\bm{v}=\bm{0}$ in equilibrium),  
subscript $0$ indicates evaluation at $r=r_0$. In the energy equation
the ratio of specific heats $\gamma$ is assumed constant, 
and we have included a thermal relaxation term with timescale $t_c$.

>>>>

We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)}
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant. 
The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = \frac{\Gamma P}{\rho}. 
\end{align}
For discussion purposes it is useful to define the vertical buoyancy frequency $N_z$ via
\begin{align}
  N_z^2 = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
and we are interested in neutrally or sub-adiabatically stratified
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

@@

The axisymmetric VGSB equations read 
\begin{align}
  & \frac{\p\rho}{\p t} + \frac{\p}{\p x}\left(\rho v_x\right) +
  \frac{\p}{\p z}\left(\rho v_z\right)=0,\\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
  v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}} = - 
  2\Omega_0\hat{\bm{z}}\times\bm{v} - S_0v_x\hat{\bm{y}} \notag\\ 
  &\phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
    v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}}=}-\frac{1}{\rho}\nabla P - \frac{\p\Phi_0}{\p z}\hat{\bm{z}},\\
  & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P = -\gamma P
  \nabla\cdot\bm{v} -
  \frac{\rho}{t_c}\left(\frac{P}{\rho}-\left.\frac{P}{\rho}\right|_{t=0}\right), 
\end{align}
where $\bm{v}$ is the velocity field relative to the bulk flow
(i.e. $\bm{v}=\bm{0}$ in equilibrium),  
subscript $0$ indicates evaluation at $r=r_0$. In the energy equation
the ratio of specific heats $\gamma$ is assumed constant, 
and we have included a thermal relaxation term with timescale
$t_c=\beta\Omega_k^{-1}$ with $\beta$  being a dimensionless input
parameter.
